---
title: "Analysis of crime data --- merge with census tracts"
subtitle: "Experiment 02"

author: "Joe Skufca"
date: "11 November 2020"
output: html_notebook
---

The goal of this analysis is to see if I can merge crime data with spatial data.

## Prepare workspace:

#### Load packages

We will work primarily within the context of the tidyverse set of packages, with a few additional packages supporting exploratory analysis.  I have included the `lubridate` package, as we will want to do some work with dates.

```{r}
library(sf)
library(leaflet)
library(tmap)
library(tmaptools)
library(tidyverse)
library(janitor)
library(readxl)
library(skimr)
library(summarytools)
library(lubridate)
```

I will also set a default theme for my ggplots.

```{r}
theme_set(theme_minimal())
```


## Data

I use two datasources in this analysis.  Spatial data (with polygons) comes via the US census data site.  

The crime data (from the crime team) has pedigree, but I do not yet have the documentation.

> TODO - document the datasource


### Read the  crime data

Andrew acquired the data and downloaded an Excel file.  The data covers the 2018 timeframe.

```{r}
thisfile="../data_raw/Crime_Incidents_in_2018.xlsx"

dfa= read_excel(thisfile) %>% clean_names()


```


#### Cleaning crime data

Data appears to be reasonably clean, though report date is character, not date.  

```{r}
dfb=dfa %>% mutate(report_dat=as_datetime(report_dat))
```


Let's also create a small dataset while we explore possiblities for prociessing.

```{r}
dfb1=dfb %>% slice_sample(n=1000)
```

### Read shapefile data


```{r}
census_sf =  st_read("../data_raw/tl_2019_11_tract/tl_2019_11_tract.shp") %>% clean_names()
```

A basic view of the census polygons.

```{r}
tmap_mode("view")
tm_shape(census_sf)+tm_polygons(alpha=.3)
```


We note that the census tract designiation is (a) not numeric and (b) not consisent with the crime data.  Let's try to find a method to create a new variable that will allow matching.

It seems that the "geoid" variable may give us a pathway to the numeric component description via the last 5 characters.

```{r}
census_sf$geoid %>% str_sub(-5,-1) %>% as.numeric() %>% sort()


dfb$census_tract %>% unique() %>% sort()

```
We will create new variable, rather than overwriting, in the census_sf dataset.

```{r}
census_sf= census_sf %>% mutate(tract_num=geoid %>% str_sub(-5,-1) %>% as.numeric())
```



## NOT FIXED BELOW


### Plotting both datasets

Create an sf object from dfb1.


```{r}
bikes1_sf = st_as_sf(dfb1, coords = c("start_lng", "start_lat"), 
                 crs = 4269, agr = "constant")
dfbu=dfb %>% mutate(rounded=round(start_lat,7)) %>%  distinct(rounded,.keep_all=T) %>% drop_na()
bikes_sf = st_as_sf(dfbu , coords = c("start_lng", "start_lat"), 
                 crs = 4269, agr = "constant")

```

```{r}

tm_shape(bikes_sf)+tm_dots(size=.01,alpha=.2)+
  tm_shape(census_sf)+tm_polygons(alpha=.3)+
  tm_basemap( leaflet::providers$OpenStreetMap )
```



### Can I do a spatial join


Leftjoin of bikes with census tract

```{r}
dfj=st_join(bikes1_sf,census_sf,join=st_within)

dfj2 =st_join(census_sf,bikes_sf,join=st_contains) %>% count(name) 

dfj2 %>% tm_shape()  +tm_polygons("n",alpha=.6)+
  tm_basemap( leaflet::providers$OpenStreetMap )

```
