---
title: "DC Income in 2018"
author: "Demetrius Boone Joe Tudor Richard Panel"
output: html_notebook
---
#Read packages
```{r}
library(sf)
library(leaflet)
library(tmap)
library(tidyverse)
library(janitor)
library(readxl)
library(skimr)
library(summarytools)
library(ggplot2)
```

Import data
```{r}
df_Income <- read.csv("../data_raw/DCincome.csv") %>% clean_names()
```

Select only the types of income we care about(Household) and the area name
```{r}
df_Income2 <- df_Income %>% select(c(geographic_area_name,estimate_households_mean_income_dollars))
```

Plot the data for visualization 
```{r}
df_Income2 %>% ggplot(aes(x=geographic_area_name, y=estimate_households_mean_income_dollars )) +
  geom_point()
```
But lets try to clean it up so you can see spefic ones you may be interested in,
first lets read shapefile data


```{r}
census_sf =  st_read("../data_raw/tl_2019_11_tract/tl_2019_11_tract.shp") %>% clean_names()
```

```{r}
tmap_mode("view")
tm_shape(census_sf)+tm_polygons(alpha=.3)
```
Rename and join files together allowing for the data to be all in one

```{r}
df_Income3 = rename(df_Income2, namelsad= geographic_area_name)

Income_sf = full_join(df_Income3, census_sf)
```

Plot to then be able to cross reference certain areas on the map with the random ones chosen to be seen.

```{r}
df_Income4= df_income3 %>% slice_sample(n=6)

df_Income4 %>% ggplot(aes(x=nameslsad, y=estimate_households_mean_income_dollars )) +
  geom_point()
```

